Gsmodutils: A python based framework for test-driven genome scale metabolic model development

### 2018-09-27 article/article.tex
More minor article changes

\author{James P Gilbert}
\email{james.gilbert@nottingham.ac.uk}
\author{Nicole Pearcy}
\author{Rupert Norman}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Thomas Millat}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Klaus Winzer}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{John King}
\affiliation[Maths]{School of Mathematical Sciences, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Charlie Hodgman}
\affiliation[Biosciences]{School of Biosciences, University of Nottingham, Sutton Bonington, Loughborough, LE12 5RD, United Kingdom}
\author{Nigel Minton}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Jamie Twycross}
\affiliation[IMA]{Intelligent Modelling and Analysis group, School of Computer Science, University of Nottingham, Nottingham, NG8 1BB, United Kingdom}

@@

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Genome scale models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady-state phenomena still provides a good understanding of system properties \cite{o2015using}.

As such approaches often leave models with missing reactions, incorrect gene-reaction rules \cite{thiele2010protocol} or with pathways based on gap filling methods that add reactions that may not be present within the organism \cite{benedict2014likelihood}.
When attempting to understand specific natural phenomena, genome annotations are frequently updated and models are often corrected in an \textit{ad hoc} manner.

@@

Therefore, models undergo significant manual annotation and curation; a process which has a high chance of error.
In this work we advocate a test-driven approach to model development highlighted in Figure \ref{fig:test_driven}.
Here, the model is changed to achieve core research goals that are dynamic in response to the changes of a project.
In order to meet this objective, validation criteria for a model such as growth conditions or the impact of gene knock outs, should be formally set.
When a model is changed, all such validation criteria should be retested to ensure that models do not regress to previous states.

@@

We feel that many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, there is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the \textit{gsmodutils} software framework.
\textit{gsmodutils} aims to provide a basis for test-driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

\subsection{Outline and features}
Test-driven development is driven by the idea of clearly defined test cases written before significant changes are made to any underlying architecture.
In the case of genome scale models, errors occur as a product of human curation to better represent newly discovered aspects of metabolism.

@@

\textit{C. autoethanogenum} is a strictly anaerobic, acetogenic bacterium which naturally produces ethanol and trace amounts of 2,3-butanediol (2,3BD) from carbon monoxide and water \cite{abrini1994clostridium,schuchmann2014autotrophy,norman2018progress}. Since carbon monoxide is readily available in the form of industrial waste gas, and 2,3BD has a global market value of \$43 billion \cite{kopke201123butanediol}, the optimisation of yields of 2,3BD from carbon monoxide is highly desirable in the context of industry \cite{norman2018progress}.   
MetaCLAU was built using Pathway Tools \cite{karp2002pathway} and ScrumPy \cite{poolman2006scrumpy}, and is based on a manually annotated genome sequence of \textit{C. autoethanogenum} \cite{humphreys2015whole}. The resulting model consisted of 758 reactions, 773 metabolites and 518 genes. For full details of the model, see (Norman et al. submitted for peer review).
The model has been integrated with the gsodutils modelling framework as a test-driven project. The following section details specific tests used to evaluate the model at each stage of its continued development.

@@

\textbf{Product scans:}
Of interest to this project were changes in the product spectrum of \textit{C. autoethanogenum} under conditions where the organism can and cannot produce molecular hydrogen (with carbon monoxide as sole carbon and energy source).
The hypothesis tested in (Norman et al., unpublished), was that in the case where hydrogen production in infeasible, alternative electron sinks like lactate and 2,3BD would be produced.
As in the previous case, the model-predicted behaviour showing both lactate and 2,3BD was deemed an important result, which model curators should be notified of if lost during model development.
Thus the analysis was built into a \textit{gsmodutils} test.

>>>>

\author{James P Gilbert}
\author{Nicole Pearcy}
\author{Rupert Norman}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Thomas Millat}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Klaus Winzer}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{John King}
\affiliation[Maths]{School of Mathematical Sciences, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Charlie Hodgman}
\affiliation[Biosciences]{School of Biosciences, University of Nottingham, Sutton Bonington, Loughborough, LE12 5RD, United Kingdom}
\author{Nigel Minton}
\affiliation[sbrc]{Synthetic Biology Research Centre, University of Nottingham, Nottingham, NG7 2RD, United Kingdom}
\author{Jamie Twycross}
\email{jamie.twycross@nottingham.ac.uk}
\affiliation[IMA]{Intelligent Modelling and Analysis group, School of Computer Science, University of Nottingham, Nottingham, NG8 1BB, United Kingdom}

@@

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Genome scale models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes or changes in cofactors and substrates for specific reactions often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady-state phenomena still provides a good understanding of system properties \cite{o2015using}.

As such, approaches often leave models with missing reactions, incorrect gene-reaction rules \cite{thiele2010protocol} or with pathways based on gap filling methods that add reactions that may not actually be catalysed by the organism in question \cite{benedict2014likelihood}.
When attempting to understand specific natural phenomena, genome annotations are frequently updated and models are often corrected in an \textit{ad hoc} manner.

@@

Therefore, models undergo significant manual annotation and curation; a process which has a high chance of error.
In this work, we advocate a test-driven approach to model development highlighted in Figure \ref{fig:test_driven}.
Here, the model is changed to achieve research goals that are dynamic in response to the changes of a project.
In order to meet this objective, validation criteria for a model such as growth conditions or the impact of gene knock outs, should be formally set.
When a model is changed, all such validation criteria should be retested to ensure that models do not regress to previous states.

@@

We feel that many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, there is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the \textit{gsmodutils} software framework.
\textit{gsmodutils} aims to provide a basis for test-driven, version controlled agile model development.
All software and packages are open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

\subsection{Outline and features}
Test-driven development is driven by the idea of clearly defined test cases written before significant changes are made to any underlying architecture.
In the case of genome scale models, errors can easily occur as a product of human curation designed to better represent newly discovered aspects of metabolism.

@@

\textit{C. autoethanogenum} is a strictly anaerobic, acetogenic bacterium which naturally produces ethanol and trace amounts of \textit{2,3-butanediol} (\textit{2,3 BD}) from carbon monoxide and water \cite{abrini1994clostridium,schuchmann2014autotrophy,norman2018progress}. Since carbon monoxide is readily available in the form of industrial waste gas, and \textit{2,3 BD} has a global market value of \$43 billion \cite{kopke201123butanediol}, the optimisation of yields of \textit{2,3 BD} from carbon monoxide is highly desirable in the context of industry \cite{norman2018progress}.   
MetaCLAU was built using Pathway Tools \cite{karp2002pathway} and ScrumPy \cite{poolman2006scrumpy}, and is based on a manually annotated genome sequence of \textit{C. autoethanogenum} \cite{humphreys2015whole}. The resulting model consisted of 758 reactions, 773 metabolites and 518 genes. For full details of the model, see (Norman et al. submitted for peer review).
The model has been integrated with the gsodutils modelling framework as a test-driven project. The following section details specific tests used to evaluate the model at each stage of its continued development.

@@

\textbf{Product scans:}
Of interest to this project were changes in the product spectrum of \textit{C. autoethanogenum} under conditions where the organism can and cannot produce molecular hydrogen (with carbon monoxide as sole carbon and energy source).
The hypothesis tested in (Norman et al., unpublished), was that in the case where hydrogen production infeasible, alternative electron sinks like lactate and \textit{2,3 BD} would be produced.
As in the previous case, the model-predicted behaviour showing both lactate and \textit{2,3 BD} was deemed an important result, which model curators should be notified of if lost during model development.
Thus the analysis was built into a \textit{gsmodutils} test.
